---
title: "HALLOWEEN MINI PROJECT"
format: html
editor: visual
---

## 

\# Define the URL for the candy data

```{r}
candy_url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

# Read the data from the URL

candy <- read.csv(url(candy_url), row.names = 1)

# Display the first few rows of the data

head(candy)
```

```{r}
# Q1: How many different candy types are in this dataset?

num_candy_types <- nrow(candy)
num_candy_types

```

```{r}
# Q2: How many fruity candy types are in the dataset?

fruity_candy_types <- table(candy$fruity)["yes"]

fruity_candy_types
```

```{r}
candy["Twix", ]$winpercent
```

```{r}
# Q3: What is your favorite candy in the dataset and what is its

candy["Sour Patch Kids", ]$winpercent
```

```{r}
#Q4. What is the winpercent value for “Kit Kat”?

candy["Kit Kat", ]$winpercent
```

```{r}
#Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

candy["Tootsie Roll Snack Bars", ]$winpercent
```


install.packages("skimr")


```{r}
library("skimr")
skim(candy)
```

Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

sugrapercnet, priccepercent and winpercent look on a different scale than the others.

Q7. What do you think a zero and one represent for the candy\$chocolate column?

A zero most likely represents "not chocolate" and a one represents "chocolate"

```{r}
# Q8: Plot a histogram of winpercent values

hist(candy$winpercent, main = "Histogram of Winpercent Values", xlab = "Winpercent")
```

Q9. Is the distribution of winpercent values symmetrical?

No it's not.

Q10. Is the center of the distribution above or below 50%?

Above %50

Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# Extract winpercent values for chocolate and fruit candies

chocolate_winpercent <- candy$winpercent[candy$chocolate == 1]

fruit_winpercent <- candy$winpercent[candy$chocolate == 0]

# Perform a t-test to compare the means

t_test_result <- t.test(chocolate_winpercent, fruit_winpercent)

# Print the t-test result

print(t_test_result)
```

chocolate candies are higher ranked than fruit candies(The mean winpercent for chocolate candies (mean of x) is 60.92153.)

Q12. Is this difference statistically significant?

Yes, the difference in winpercent values between chocolate and fruit candies is statistically significant. This is indicated by the very small p-value (4.164e-10), which is well below the typical significance level of 0.05.

```{r}
# Q13. What are the five least liked candy types in this set?

least_liked <- head(candy[order(candy$winpercent), ], n = 5)

least_liked


```

```{r}
# Q14. What are the top 5 all-time favorite candy types out of this set?

favorite <- head(candy[order(candy$winpercent, decreasing = TRUE), ], n = 5)

favorite
```

```{r}
library(ggplot2)
```

```{r}
ggplot(candy, aes(x = winpercent, y = reorder(row.names(candy), winpercent))) +

  geom_bar(stat = "identity", fill = "skyblue") +

  labs(title = "Candy Ranking Based on Winpercent Values",

       x = "Winpercent", y = "Candy") +

  theme_minimal()
```

```{r}
my_cols=rep("black", nrow(candy))

my_cols[as.logical(candy$chocolate)] = "chocolate"

my_cols[as.logical(candy$bar)] = "brown"

my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 

  aes(winpercent, reorder(rownames(candy),winpercent)) +

  geom_col(fill=my_cols) 
```

Q17. What is the worst ranked chocolate candy?

sixlets

\- Q18. What is the best ranked fruity candy?

starburst

install.packages("ggrepel")

```{r}
library(ggrepel)
```

```{r}
library(ggrepel)
```

```{r}
# How about a plot of price vs win

ggplot(candy) +

  aes(winpercent, pricepercent, label=rownames(candy)) +

  geom_point(col=my_cols) + 

  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)

head( candy[ord,c(11,12)], n=5 )
```

Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Hershey's Milk Chocolate

Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip

install.packages("corrplot")

```{r}

library(corrplot)
```

```{r}
cij <- cor(candy)

corrplot(cij)
```

Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

winpercent and pricepercent

Q23. Similarly, what two variables are most positively correlated?

chocolate and fruity

```{r}
pca <- prcomp(candy, scale = TRUE)

summary(pca)
```

```{r}
plot(pca$x[, 1], pca$x[, 2], xlab = "PC1", ylab = "PC2")
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data

my_data <- cbind(candy, pca$x[,1:3])
```
install.packages("ggplot2")

```{r}

library(ggplot2)
```

```{r}
p <- ggplot(my_data) + 

        aes(x=PC1, y=PC2, 

            size=winpercent/100,  

            text=rownames(my_data),

            label=rownames(my_data)) +

        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)
```

```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 

  theme(legend.position = "none") +

  labs(title="Halloween Candy PCA Space",

       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",

       caption="Data from 538")
```


install.packages("plotly")


```{r}
library(plotly)
```

```{r}
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))

barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity candies are picked up by PC1 in the positive direction. This doesn't make sense to me. The positive association of fruity candies with PC1 suggests that candies with fruity characteristics are more likely to have higher winpercent values, indicating higher popularity, but winpercent and fruity has the opposite effects.
