---
title: "Machine learning1"
author: "Shahriyar Chamankhah"
date: "2024-05-01"
output: html_document
---

# First up kmeans()

Demo of using kmeans() function in base R. first make up some data with a known structure.

```{r}
tmp <- c( rnorm(30, -30), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))
x
plot(x)
```

Now we have soem made up data in 'x' let's see how kmeans works with this data


```{r}
k <- kmeans(x, centers = 2, nstart = 20)
k
```

Q. How many points are in each cluster

```{r}
k$size
```

Q. How we do get to the cluster membership/assignment.

```{r}
k$cluster
```

Q. What about cluster centers?

```{r}
k$centers
```

Now we got to the main results let's use them to plot our data with the kmeans result.

```{r}
plot(x, col=k$cluster)
points(k$centers, col="Blue", pch=15)
```

## Now for hclust(

we will cluster the same data 'x' with the hclust()
in this case hclust() requires a distance matrix as input

```{r}
hc <- hclust(dist(x))
hc

```

```{r}
plot(hc)
```

To get our cluster membership vector we need to "cut" the tree with the 'cutree()'

```{r}
grps <- cutree(hc, h=8)
grps
```


ow plot our data with the hclust() results.

```{r}
plot(x, col=grps)
```





#Principal Component Analysis(PCA)
##PCA of UK food data


Read data from website and try a few visualizations.

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names = 1)
x
```


```{r}
cols <- rainbow(nrow(x))
barplot(as.matrix(x),col = cols)
```

```{r}
barplot(as.matrix(x),col = cols, beside = TRUE)
```


```{r}
pairs(x, col=cols)
```


PCA to the rescue
the main base R PCA fucntion is called 'prcomp()'
and we will need to give it the transpose of our input data



```{r}
pca <- prcomp(t(x))


```


There is a nice summary of how well PCA is doing

```{r}
summary(pca)

```

```{r}
attributes(pca)
```

To make our new PCA plot(a.k.a PCA score plot) we access 'pca$x

```{r}
plot(pca$x[,1], pca$x[,2])
text(pca$x[,1], pca$x[,2], colnames(x))
```

color up  the plot


```{r}
country_cols <- c("orange", "red", "blue", "green")
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab = "PC2")
text(pca$x[,1], pca$x[,2], colnames(x), col=country_cols)
```

## PCA of RNA-SEQ data

Read in data from website

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```



```{r}
pca <- prcomp(t(rna.data)) 
summary(pca)
```
 Do the plot
 
```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
```
 



```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
text(pca$x[,1], pca$x[,2], colnames(rna.data))
```
